Grove: A Bidirectionally Typed Collaborative Structure Editor Calculus

### 2024-10-25 review/response.md

Thank you for your careful readings and thoughtful comments! We are heartened by the encouraging comments and strong overall assessment from RA and RB.

We do not believe RC has identified any significant technical deficiencies in the work. We hope the clarifications we provide below will help illuminate some subtleties regarding our claimed contributions and the technical developments in the paper. We hope that this will lead to a more positive final assessment from RC as well.

# Review A

We agree with the detailed comments and will address them in the final version of the paper.

# Review B

> My only (very minor) complaint is that as somewhat of an outsider to this space, I would have appreciated a longer and more detailed related works section, just to understand more about how the core ideas fit into the broader space. This is not a demand for a longer Related Works–I think the section is clearly already sufficient for acceptance. I only mention this because I think the paper could make a stronger case for its core ideas by offering a deeper analysis of the research space, especially for readers who are not already familiar with the relevant areas.

We agree, and we will fortify our discussion of related works in the final version of the paper. We intend to add further discussion of the literature on CRDTs and on patch synthesis approaches in version control systems in particular.

# Review C

> For the first 18 pages, the paper develops a graph representation for representing
edits and various kinds of conflicts.

It is indeed the case that the central technical idea of the paper is a novel graph representation of collaboratively edited programs (L104), a corresponding patch calculus with commutative edits (L110), and a corresponding invertible decomposition into a grove (L116).

> The graph representation is essentially a different way of representing textual information by using special nodes and edge/node relations to represent conflicts.

Note that the graph representation captures information that cannot be represented in ordinary program text - namely, a subterm occurring in multiple places, multiple subterms in the same location, and a term containing itself as a subterm. To this end, the graph includes unique identifiers of vertices and edges, which are not contained in a typical textual representation. Our decomposition of the graph into groves, which may be presented "textually" to the user as in our implementation, was carefully structured to retain this information, as we prove in Agda.

> Compared to Hazelnut and Hazel, the only difference is that this paper assigns all conflicts to have the Dyn type from gradual typing. ... The idea of using Dyn types were already introduced in Hazel, so there is very little novelty here.

Section 5 on type checking was intended to be a "cherry on top" of the main contributions listed above. It is indeed the case that the type system is based directly on the marked lambda calculus, which is the core calculus of Hazel. However, the contribution of Section 5 is a bit more nuanced than is stated in this review:

- Local conflicts in synthetic position synthesize the dynamic type and generate constraints between this type and the synthesized types of the contents.
- Relocation and unicycle conflicts in synthetic position synthesize the dynamic type with provenance determined by the identifier they refer to. 
- Local conflicts in analytic positions analyze their contents against the same analyzed type.
- Relocation and unicycle conflicts in analytic position generate a constraint between the analyzed type and the dynamic type with provenance determined by the identifier they refer to.

These constraints allow the integration of typing information from all references to the same identifier.

> The paper keeps on saying that "Grove does not suffer from problems due to merge conflicts". However, the only reason this is true is that Grove represents/encapsulates these conflicts. It does not solve any of the merging problems.

The paper does not claim that "Grove does not suffer from problems due to merge conflicts." On line 87, it claims that Grove "does not suffer from the problems just outlined," those problems being the granularity problem (L46), the relocation conflict problem (L52), and the relocation modification problem (L56). Each problem is addressed in Grove as described in Sec. 2.

The paper also states that "The Grove patch language is commutative, meaning that there is no need for a complex three-way merge algorithm (i.e. operational transform)" (L288). Again, this is not a claim that there are no problems due to merge conflicts, but rather that we do not need to rely on that particularly complex kind of algorithm to discover the merge conflicts.

We can see how these statements might not be as clear as possible in emphasizing that Grove does indeed represent/encapsulate merge conflicts without resolving them on the user's behalf fully automatically, and will expand on this in both the introduction and conclusion of the final version of the paper so that readers are less likely to misunderstand the claim.

We believe that automatically resolving merge conflicts would require ad-hoc heuristics and inhibit user agency. Instead, Grove provides a setting in which those unresolved merge conflicts are syntactically and semantically meaningful. In particular, conflicts do not block edits from being made and shared with collaborators, and conflicts do not block the program's semantics. Previous version control systems require merge conflicts to be resolved before editor services may be available again, and Grove eliminates this gap problem by supporting editor services at all times.

Automatic heuristics could be layered on top of Grove as a convenience, and indeed we mention a few possible situations where this might be helpful in the paper (e.g. in the paragraph on L390 and the next paragraph on L415).

Details:

> Section 2.3.2 It is well known that tree-based editors handle the granularity problem better than line-based. But the tree-based editor has a much higher complexity, building the tree and maintaining the correct tree structure during edits such as cut and paste.

Hazel is a structured editor, so the tree structure of the program is maintained natively in the editor. It's true that this is more complex than just maintaining a string, but this complexity is accepted and handled in other aspects of Hazel's theory and implementation (e.g. Omar et al., 2017, Moon et al., 2023). There are a number of successful structure editors in popular usage, e.g. Scratch.

> Section 2.3.3 Grove chooses to silently merge changes from relocation modification, which current versioning control systems often yield a merge conflict. However, it is hard to say that this behavior is always wanted or preferred. In many scenarios, the involved programmers may want to review before such merges are committed. Also, this example seems to work well because the operations involved, + and *, are commutative. What if we have operations like - and /?

The key point here is that Alice's and Bob's edits are not in conflict with each other - one is modifying an expression, and the other is modifying the location of that expression. We posit that in practice, such actions are morally independent, even though they affect some of the same parts of the tree. We acknowledge, however, that in some cases like these it may be helpful to notify the users of the proximity of the edits. Such a notification system would not affect the underlying theory. We will note this possibility in the next version of the submission.

When two edits are truly in conflict, meaning there is no way to apply them both and maintain a correct syntax tree, then Grove 'merges' them, but not silently - it results in a syntactically marked conflict.

The commutativity of the operator here does not matter -- Grove operates purely syntactically.

> Figure 10(d), why not add an L edge from +8 to *2 and a R edge from +8 to *24? Also, why R31 and not R35 and L33 not L37?

The R edge from +8 to *24 is not kept in 10(d) for the same reason that it is not kept in 10(b) - because Alice's action is to relocate *24, which involves both adding an edge and deleting an edge. When this same action is played back on 10(c), the R27 edge is deleted. Playing back all of the merged actions ensures commutativity. The edge names are different for a similar reason - the actions depicted delete and insert edges, so the final edges are new and have new identifiers.

> What is the connection between Line 367 and Line 531? Earlier it says a disconnected term may be connected back after a deletion but later it says deleted edges can't connect again. Also, since an edge includes a term, the deletion of the edge also causes the deletion of the term if it is connected to by only one edge.

Deleting an edge does not delete any vertices. A term may be disconnected by deleting an edge and may be reconnected by inserting a new edge with the same origin and destination but a different identifier. The old edge remains deleted, but the term is reconnected. An edge does not include a term - it refers to an origin location and destination vertex.

> In Rule MKSLocalConflict, it is possible to synthesize a more precise type than ?. A more meaningful choice is the least upper bound of all σi's in terms of some subtyping relation. For example, if all branches of a conflict have Int, then it is fine to say that the conflict has the type Int.

This is true, and we considered taking this approach. We decided not to because it is more complete and modular to defer to the type hole inference system by emitting constraints that each synthesized type from the contents must be equal to the synthesized gradual type (and hence to each other). This allows the type hole inference system to indeed infer Int if they all have type Int, as well as to infer more complete information than the least upper bound would.

> In rules MALocalConflict, MARelationConflict, and MAUnicycleConflict, why do you generate a constraint between a ? and σ? In gradual typing, such constraints are simply dropped. Probably you want to use the location for provenance. However, it is unclear how the location associated with each ? is used.

Indeed, such constraints are not dropped, and indeed we use the location for provenance. This section extends the work on the marked lambda calculus and type hole inference presented in Zhao et al., 2024. This prior work describes the details of how the provenances and constraints are solved and used. To summarize, constraints are solved with each instance of the gradual type from a specific provenance as a metavariable. This allows constraints on the same provenance originating at different points in the bidirectional flow to be combined to form a solution. If solved, this solution is returned to the bidirectional typing flow, and if conflicts are found, the possible solutions for type holes in the surface syntax are offered to the user as an editing affordance.

As it happens, the constraint generated in the rule MALocalConflict is actually redundant because the hole with provenance $\exp(\ell)$ cannot appear in any other constraints generated by the program. This unnecessary constraint will be removed in the final version of the paper. We will also streamline the rules by altering MSLocalConflict to analyze the contents against the synthesized gradual type rather than generating constraints directly. Neither of these is an issue of correctness, but each represents an improvement to the presentation.

Typos:

> Line 163, what does "this work" refer to? Probably it refers to Hazelnut and Hazel, but it is good to be explicit

"This work" refers to "the previous work on the Hazelnut structure editor calculus" (L139), and more specifically, "a type system and type error localization system (the marked lambda calculus) for incomplete programs" (L140) mentioned in the previous paragraph. We will make this reference more explicit in the next version of the submission.

> Lemma 3.8, I am not sure if this is actually correct. For example \*24 in Figure 9(d) is both an MP and a U root.

Lemma 3.8 is correct. \*24 in Figure 9(d) is an MP root, but it is not a U root because there is no nontrivial sequence of vertices that begins and ends at \*24 such that each vertex is the *unique* parent of the previous. \*24 does not have a unique parent - it has multiple parents.

### 2024-10-25 OriginalSubmissionSource/abstract.tex
Updates from Overleaf

\begin{abstract}
Version control systems typically rely on a \emph{patch language}, heuristic \emph{patch synthesis algorithms} like \li{diff}, and \emph{three-way merge algorithms}. Standard patch languages and merge algorithms often fail to identify conflicts correctly when there are multiple edits to one line of code or code is relocated. This paper introduces Grove, a collaborative structure editor calculus  that eliminates patch synthesis and three-way merge algorithms entirely. 
Instead, patches are derived directly from the log of the developer's edit actions 
and all edits commute, i.e. the repository state forms a commutative replicated data type (CmRDT). To handle conflicts that can arise due to code relocation, 
the core datatype in Grove is a labeled directed multi-graph with uniquely identified 
vertices and edges. All edits amount to edge insertion and deletion, with deletion being permanent. 
To support tree-based editing, we define a decomposition from graphs into \emph{groves}, which are a set of syntax trees with conflicts---including multi-child, multi-parent, and cycle conflicts---represented explicitly using holes and references between trees. Finally, we define a type error localization system for groves that enjoys a \emph{totality} property, i.e. all editor states in Grove are statically meaningful, so developers can use standard editor services while 
working to resolve these explicitly represented conflicts. The static semantics is defined as a bidirectional marking system in line with recent work, with gradual typing employed to handle situations where errors and conflicts 
prevent type determination. We then layer on a unification-based type inference system  to opportunistically fill type holes and fail gracefully when no solution exists. We mechanize the metatheory of Grove using the Agda theorem prover. We  implement these ideas as the \emph{Grove Workbench}, which generates the necessary data structures and algorithms in OCaml given a syntax tree specification.
\end{abstract}

>>>>

\begin{abstract}
Version control systems typically rely on a \emph{patch language}, heuristic \emph{patch synthesis algorithms} like \li{diff}, and \emph{three-way merge algorithms}. Standard patch languages and merge algorithms often fail to identify conflicts correctly when there are multiple edits to one line of code or code is relocated. This paper introduces Grove, a collaborative structure editor calculus  that eliminates patch synthesis and three-way merge algorithms entirely. 
Instead, patches are derived directly from the log of the developer's edit actions 
and all edits commute, i.e. the repository state forms a commutative replicated data type (CmRDT). To handle conflicts that can arise due to code relocation, 
the core datatype in Grove is a labeled directed multi-graph with uniquely identified 
vertices and edges. All edits amount to edge insertion and deletion, with deletion being permanent. 
To support tree-based editing, we define a decomposition from graphs into \emph{groves}, which are a set of syntax trees with conflicts---including local, relocation, and unicyclic relocation conflicts--- represented explicitly using holes and references between trees. Finally, we define a type error localization system for groves that enjoys a \emph{totality} property, i.e. all editor states in Grove are statically meaningful, so developers can use standard editor services while 
working to resolve these explicitly represented conflicts. The static semantics is defined as a bidirectional marking system in line with recent work, with gradual typing employed to handle situations where errors and conflicts 
prevent type determination. We then layer on a unification-based type inference system  to opportunistically fill type holes and fail gracefully when no solution exists. We mechanize the metatheory of Grove using the Agda theorem prover. We  implement these ideas as the \emph{Grove Workbench}, which generates the necessary data structures and algorithms in OCaml given a syntax tree specification.
\end{abstract}

### 2024-10-25 OriginalSubmissionSource/grove-by-example.tex
Updates from Overleaf

Grove can form the basis for both a conventional version control workflow,
where edits are batched into commits, or real-time collaborative editing, 
where edits are communicated as they occur. This paper makes no assumptions about which batching mode is in use, nor do we consider the well-studied problem of reliably and efficiently communicating patches over a network\todo{cite something, dist sys textbook?}.\todo{maybe include a brief note in the discussion about cursors being represented by locations?} 
%\autoref{sub:Cursors} discusses representing cursor locations when in real-time mode.

@@

Each vertex represents a term in the specified language, except for a distinguished root vertex, 
and is labeled with a unique identifier (UID) and a \emph{constructor}. 
In \autoref{fig:Simple:a}, the root vertex has UID 0 and constructor~\textbullet. 
Vertices \vSimpleTimes{} and \vSimpleX{} have UIDs 2 and 4 and constructors~\texttt{*} and \texttt{var(}$x$\texttt{)}, respectively.
For clarity, we abbreviate \texttt{var(}$x$\texttt{)} as simply \texttt{x}; here, $x$ is a constructor parameter. We treat identifiers and literals as indivisible, but we discuss character-level editing in Section~TODO:REF\todo{ref}.

@@

For clarity, we use odd numbers for vertex UIDs and even numbers for edge UIDs. In practice, UIDs would be generated by a mechanism
that effectively ensures that collaborators always generate distinct UIDs, e.g. by generating
universally unique IDs (UUIDs)~\cite{IETF RFC 4122}.

@@

\subsection{Edits}%
\label{sub:Single-User Actions}
\figureWrapMove

Individual users perform \emph{edits} to evolve the edit state. This paper abstracts over the user interface aspects of structure editors and makes no usability-related claims; edits could be performed through, for example, drag-and-drop interactions (as in block-based editors like Scratch) or keyboard interactions (as in MPS and Hazel). % User interface design for structure editors remains an active research area.
% The common thread is that structure editors maintain a program sketch throughout the editing process.

@@

We now turn our attention to how Grove handles collaboration.
The examples in this section generalize to collaborations between any number of users,
but for simplicity we consider only two: Alice and Bob.
Alice and Bob are each concurrently editing their own branches of the repository (or their own instance of a real-time collaborative editor), 
performing edits that translate to patches as described above. 
They periodically communicate these patches to one another. \autoref{fig:Commutativity} diagrams the Grove collaboration model.\todo{stretch: include a comparable diagram for }

@@

Alice and Bob start where Alice left off in \autoref{fig:Move:b} 
with the term \texttt{\hole{} + \hole{} * y}.
Alice then adds~\vDifferentPartsAlice{} as the left child of \vSimpleTimes{}.
Concurrently, Bob changes \vSimpleY{} to \vDifferentPartsBob{}.
Before sharing their patches,
Alice and Bob have the edit states \autoref{fig:DifferentParts:a} and \autoref{fig:DifferentParts:b}, respectively.
Note that the transition from \autoref{fig:Move:b} to \autoref{fig:DifferentParts:b}
represents multiple graph updates,
i.e., deleting~\eSimpleY{} and adding~\eDifferentPartsBob{} along with its child~\vDifferentPartsBob{}.
We thus mark the transition with a star.
Once Alice and Bob share their patches and apply each other's patch to their own edit state,
both edit states converge to the graph in \autoref{fig:DifferentParts:c}.\todo{add more spacing around terms in Fig 5} 
Because Grove's patch language is structural rather than line-based, the fact that these edits happened to be close to one another (i.e. in the same arithmetic expression) does not run afoul of the \textbf{granularity problem} described in \autoref{sec:Introduction}.

@@

After converging on \texttt{\hole{} + u * y} in \autoref{fig:DifferentParts:c}, 
Alice changes~\vDifferentPartsAlice{} to~\vNestedPartsAlice{},
producing the edit state in \autoref{fig:NestedParts:a}.
Meanwhile, Bob relocates~\vSimpleTimes{} (bringing along its children)
from the~\texttt{R} position of~\vWrapPlus{} to the~\texttt{L} position. As discussed above, this involves deleting~\eMoveTimes{} and adding~\eNestedPartsBob{}.
Bob's resulting edit state is shown in \autoref{fig:NestedParts:b}.\todo{}

@@

The collaborative edits discussed so far merge cleanly,
but in general, merging patches can lead to graphs that do not map cleanly to a conventional syntax tree. We identify several different motifs that might arise, all of which give rise to different kinds of conflicts in the graph decomposition: \emph{local conflicts}, \emph{multi-location conflicts}, and \emph{cyclic location conflicts}. 
As with merge conflicts in version-control systems such as git, these all require user intervention to resolve.

@@

When these patches are merged, \emph{both} \eMultiChildAlice{} and \eMultiChildBob{}
appear in the merged graph. 
When decomposing this graph to a syntax tree, 
we resolve this conflict in the~\texttt{R} position of~\vWrapPlus~by decomposing to a \emph{local conflict}, \figureCode{$\binaryConflictHole{x}{y}$}, as 
shown in \autoref{fig:MultiChild:c}.

@@

Such conflicts can be resolved simply by deleting relocating one of the conflicting terms (and editing the remaining term into the correctly merged value, if needed), which would remove the corresponding edge.
For example, Alice could resolve the problem by wrapping \li{x} and \li{y} with a multiplication, effectively moving them to non-conflicting locations.
No special edit actions are needed for conflict resolution.

@@

It is worth noting one special situation: if Alice and Bob both independently filled the hole with structurally identical terms, e.g. \li{x}, Grove would \emph{still} formally identify a conflict, because each \li{x} has a distinct UID. In this situation, it would be reasonable for the system to resolve the conflict without further coordination by deterministically choosing one of the two UIDs, e.g. the smallest one.

@@

As we will return to in \autoref{sec:Type System}\todo{sec}, 
a local conflict need not disable type-based editor services (the \textbf{semantic gap problem}). Instead, the system can type check around the conflict (and if the conflicted terms have a common type, the system can infer that type for the location conflict itself).

\subsubsection{Multi-Location Conflicts}%
\label{sub:Multi-parent conflicts}

@@

Multi-parent conflicts occur when multiple edges share a destination vertex,
as opposed to multi-child conflicts, which share a common origin.
An example of this is in \autoref{fig:MultiParent},
which shows the edit state Alice and Bob converge on after deleting the conflict and the \texttt{v}.
Alice then moves~\vNestedPartsAlice{} to the~\texttt{R} position of~\vSimpleTimes{}.
At the same time, Bob also moves~\vNestedPartsAlice{} to the~\texttt{R} position of~\vWrapPlus{}.
This results in the edit states of \autoref{fig:MultiParent:b} and \autoref{fig:MultiParent:c}
In both cases, the edits are modeled as one edge deletion followed by one edge addition.
Alice deletes~\eNestedPartsAlice{} and adds~\eMultiParentAlice{}.
At the same time, Bob deletes~\eNestedPartsAlice{} and adds~\eMultiParentBob{}.
As we explain in \autoref{sec:Formalism}, deletions are idempotent,
so the fact that Alice and Bob both deleted~\eNestedPartsAlice{} will not lead to a conflict.
However, both~\eMultiParentAlice{} (added by Alice) and~\eMultiParentBob{} (added by Bob)
point to the same vertex.
Once all updates are merged, the resulting edit state is given in \autoref{fig:MultiParent:d}.
Notice~\vNestedPartsAlice{} has two edges pointing to it.

TODO: talk about how to resolve the conflict

@@

\figureCycle

In addition to multi-child and multi-parent conflicts, merges
can cause cycles in the graph.\footnote{In theory a single user could
generate cycles, but this could be detected immediately and the user action
that caused that cycle rejected.
Thus, we are only concerned with cycles that cannot be immediately detected
and prevented by the editor.}
For example, consider the situation in \autoref{fig:Cycle}.
\autoref{fig:Cycle:a} shows the graph after Alice and Bob have made some edits,
and shared them with each other.
Alice then moves~\vMultiCycleTimes{} to the \texttt{R} child of~\vWrapPlus{}
and then~\vMultiCyclePlus{} underneath that.
Bob does the same but puts~\vMultiCycleTimes{} under~\vMultiCyclePlus{}.
This results in \autoref{fig:Cycle:b} and \autoref{fig:Cycle:c}, respectively.

On their own, neither of these edits creates a cycle.
However, merging the edit actions of both Alice and Bob results in the graph
in \autoref{fig:Cycle:d}, which has a cycle
between~\vMultiCycleTimes{} and~\vMultiCyclePlus{}.
As with the other kinds of conflict, resolving it requires user input.
For example, by either deleting both~\eMultiCycleBobPlus{} and~\eMultiCycleBobTimes{}~(thus favoring Alice's version)
or deleting both~\eMultiCycleAliceTimes{} and~\eMultiCycleAlicePlus{}~(thus favoring Bob's version).

\paragraph{Disconnected Cycles}

\figureDisconnect

@@

Finally, there is a particular type of cycle worth discussing.

Since a cycle like this that is connected to the root vertex always
contains vertexes with multiple parents that if removed would break the cycle,
we use the user interface described in \autoref{sub:Multi-parent conflicts}
to display these kinds of programs.
For example, \autoref{fig:Cycle:d} would be displayed.
We leave the user interface considerations of resolving the conflict such a
cycle represents to future work.

Finally, when multi-user edits are merged,
parts of the graph can become disconnected from the root
even though they are connected to the root in each user's
copy of the graph before the merge.

For example, consider the situation in \autoref{fig:Disconnect:a},
which shows the graph after Alice and Bob have made some edits,
and shared their edits with each other.
Alice then moves~\vMultiCycleTimes{} under~\vSimpleTimes{},
while Bob moves~\vSimpleTimes{} under~\vMultiCycleTimes{}.
This results in \autoref{fig:Disconnect:b} and \autoref{fig:Disconnect:c}, respectively.

When these edits are merged, the result is \autoref{fig:Disconnect:d}.
Since Alice deleted~\eNestedPartsBob{}
and Bob deleted~\eMultiCycleAliceTimes{},
when these are merged~\vMultiCycleTimes{} and~\vSimpleTimes{}
are disconnected from the root.

This is similar to the type of merge conflict
that can arise in traditional version control systems
when two users delete different functions that
are duplicates of the other.

Fortunately, we can detect this situation and warn the user
since the disconnection appears after merging two graphs
in which vertexes were not already disconnected.

>>>>

Grove can form the basis for both a conventional version control workflow,
where edits are batched into commits, or real-time collaborative editing, 
where edits are communicated as they occur. This paper makes no assumptions about which batching mode is in use, nor do we consider the well-studied problem of reliably and efficiently communicating patches over a network.
%\autoref{sub:Cursors} discusses representing cursor locations when in real-time mode.

@@

Each vertex represents a term in the specified language, except for a distinguished root vertex, 
and is labeled with a unique identifier (UID) and a \emph{constructor}. 
In \autoref{fig:Simple:a}, the root vertex has UID 0 and constructor~\textbullet. 
Vertices \vSimpleTimes{} and \vSimpleX{} have UIDs 2 and 4 and constructors~\texttt{*} and \texttt{var(}$x$\texttt{)}, respectively.
For clarity, we abbreviate \texttt{var(}$x$\texttt{)} as simply \texttt{x}; here, $x$ is a constructor parameter. We treat identifiers and literals as indivisible, but we discuss character-level editing in \autoref{sec:Discussion and Conclusion}.

@@

For clarity, we use odd numbers for vertex UIDs and even numbers for edge UIDs. In practice, UIDs would be generated by a mechanism
that effectively ensures that collaborators always generate distinct UIDs, e.g. by generating
universally unique IDs (UUIDs)~\cite{paskin1999toward}.

@@

\subsection{Structure Editing}%
\label{sub:Single-User Actions}
\figureWrapMove

Individual users perform \emph{edits} to evolve the edit state. We consider several standard edits, including insertion, deletion, cut-and-paste (relocation), copy-and-paste, and undo/redo. This paper abstracts over the user interface aspects of structure editors and makes no usability-related claims; these edits could be performed through, for example, drag-and-drop interactions (as in block-based editors like Scratch) or keyboard interactions (as in MPS and Hazel).  % User interface design for structure editors remains an active research area.
% The common thread is that structure editors maintain a program sketch throughout the editing process.

@@

We now turn our attention to how Grove handles collaboration.
The examples in this section generalize to collaborations between any number of users,
but for simplicity we consider only two: Alice and Bob.
Alice and Bob are each concurrently editing their own branches of the repository (or their own instance of a real-time collaborative editor), 
performing edits that translate to patches as described above. 
They periodically communicate these patches to one another. \autoref{fig:Commutativity} diagrams the Grove collaboration model.

@@

Alice and Bob start where Alice left off in \autoref{fig:Move:b} 
with the term \texttt{\hole{} + \hole{} * y}.
Alice then adds~\vDifferentPartsAlice{} as the left child of \vSimpleTimes{}.
Concurrently, Bob changes \vSimpleY{} to \vDifferentPartsBob{}.
Before sharing their patches,
Alice and Bob have the edit states \autoref{fig:DifferentParts:a} and \autoref{fig:DifferentParts:b}, respectively.
Note that the transition from \autoref{fig:Move:b} to \autoref{fig:DifferentParts:b}
represents multiple graph updates,
i.e., deleting~\eSimpleY{} and adding~\eDifferentPartsBob{} along with its child~\vDifferentPartsBob{}.
We thus mark the transition with a star.
Once Alice and Bob share their patches and apply each other's patch to their own edit state,
both edit states converge to the graph in \autoref{fig:DifferentParts:c}. 
Because Grove's patch language is structural rather than line-based, the fact that these edits happened to be close to one another (i.e. in the same arithmetic expression) does not run afoul of the \textbf{granularity problem} described in \autoref{sec:Introduction}.

@@

After converging on \texttt{\hole{} + u * y} in \autoref{fig:DifferentParts:c}, 
Alice changes~\vDifferentPartsAlice{} to~\vNestedPartsAlice{},
producing the edit state in \autoref{fig:NestedParts:a}.
Meanwhile, Bob relocates~\vSimpleTimes{} (bringing along its children)
from the~\texttt{R} position of~\vWrapPlus{} to the~\texttt{L} position. As discussed above, this involves deleting~\eMoveTimes{} and adding~\eNestedPartsBob{}.
Bob's resulting edit state is shown in \autoref{fig:NestedParts:b}.

@@

The collaborative edits discussed so far merge cleanly,
but in general, merging patches can lead to graphs that do not map cleanly to a conventional syntax tree. We identify several different motifs that might arise, all of which give rise to different kinds of conflicts in the graph decomposition: \emph{local conflicts}, \emph{relocation conflicts}, and \emph{unicyclic relocation conflicts}. 
As with merge conflicts in version-control systems such as git, these all require user intervention to resolve. 
%As we will return to in \autoref{sec:Type System}\todo{sec}, conflicts need not disable type-based editor services (the \textbf{semantic gap problem}). Instead, the system can type check around these conflict (and in some cases, infer a type for the conflict itself).

@@

When these patches are merged in \autoref{fig:MultiChild:c}, \emph{both} \eMultiChildAlice{} and \eMultiChildBob{}
appear in the merged graph. 
When decomposing this graph to a syntax tree, 
we resolve this conflict in the~\texttt{R} position of~\vWrapPlus~by decomposing to a \emph{local conflict}, {\binaryConflictHole{x}{y}}.

@@

Local conflicts can be resolved simply by deleting or relocating all but one of the conflicting terms (and editing the remaining term into the correctly merged value, if needed), which would remove the corresponding edge.
For example, Alice could resolve the problem by wrapping \li{x} and \li{y} with a multiplication, effectively moving them to non-conflicting locations.
No special edit actions are needed for conflict resolution.

@@

It is worth noting one special situation: if Alice and Bob independently filled the hole with structurally identical terms, e.g. \li{x}, Grove would \emph{still} formally identify a conflict, because the terms have distinct UIDs. In this situation, it would be reasonable for the system to resolve the conflict without further coordination by deterministically choosing one of the two terms, e.g. the smallest.

@@

\subsubsection{Relocation Conflicts}%
\label{sub:Multi-parent conflicts}

@@

Grove's support for repositioning creates the possibility for \emph{relocation conflicts}. These occur when a merge causes a vertex to have multiple incoming edges, indicating that it does not have a uniquely determined location (as opposed to local conflicts, which occur when there are multiple outgoing edges at a specified location).

For example, \autoref{fig:MultiParent} shows an example where Alice and Bob relocate a term, \li{w}, to two different locations (the two holes in \autoref{fig:MultiParent:a}). In both cases, the edits are modeled as one edge deletion followed by one edge addition.
Alice deletes~\eNestedPartsAlice{} and adds~\eMultiParentAlice{} in \autoref{fig:MultiParent:b}.
At the same time, Bob deletes~\eNestedPartsAlice{} and adds~\eMultiParentBob{} in \autoref{fig:MultiParent:c}.
Edge deletions is idempotent,
so the fact that Alice and Bob both deleted~\eNestedPartsAlice{} will not lead to a conflict. 
However, both~\eMultiParentAlice{} (added by Alice) and~\eMultiParentBob{} (added by Bob)
point to the same vertex.
Once these patches are merged, the resulting edit state is given in \autoref{fig:MultiParent:d}.
Notice~\vNestedPartsAlice{} has two edges pointing to it. 
When decomposing the graph, we leave a \emph{relocation conflict reference}, $\multiVertex{18}$, at each conflicting location. The conflicted term is separately tracked in the result of decomposition, which, due to conflicts like these, is formally a set of terms with references like these between them. We call this set a \emph{grove}. 
This approach partially addresses the \textbf{relocation conflict problem} from \autoref{sec:Introduction} (we also need to handle unicycles, see below, to fully address the problem).

To resolve relocation conflicts, a user can simply delete all but one of the relocation conflict references. This will cause the corresponding edges to be deleted,
and when only one edge remains, there will no longer be a conflict. An editor might provide a more convenient way of deleting all but a selected relocation conflict reference, and provide affordances for displaying these terms, e.g. by transcluding them inline at each location or showing them in a separate sidebar.

@@

\figureCycle
\figureDisconnect

Relocation in a collaborative setting can also lead to cycles in the graph. 
For example, consider the situation in \autoref{fig:Cycle}.
Starting in \autoref{fig:Cycle:a}, 
Alice relocates~\vMultiCycleTimes{} to the \texttt{R} child of~\vWrapPlus{}
and then~\vMultiCyclePlus{} underneath that in \autoref{fig:Cycle:b}.
Bob does the opposite, putting~\vMultiCycleTimes{} under~\vMultiCyclePlus{} in \autoref{fig:Cycle:c}. On their own, neither of these edits creates a cycle.
However, merging the two patches results in the graph
in \autoref{fig:Cycle:d}, which has a cycle
between~\vMultiCycleTimes{} and~\vMultiCyclePlus{}.

The main difficulty with cycles has to do with decomposition back to a term, i.e. a syntax tree. We do not want decomposition to traverse endlessly attempting to create an infinite tree, so we need to break the cycle somewhere.

If the cycle is connected to a larger term, then there will necessarily be at least one vertex along the cycle that has multiple incoming edges. In \autoref{fig:Cycle}, both \vMultiCycleTimes{} and \vMultiCyclePlus{} have multiple incoming edges. As described above, decomposition will leave relocation conflict references in these positions, thereby breaking the cycle. In this example, these references appear within a local conflict as well, because both vertices were relocated under a common parent vertex. As before, this cycle can be broken by deleting or otherwise modifying the terms until there are no longer any such conflicts.

It is also possible to merge patches such that a \emph{disconnected unicycle} emerges in the graph, even when neither patch disconnects any vertex from the root. \autoref{fig:Disconnection} shows a simple example of when this could occur: Alice relocates~\vMultiCycleTimes{} under~\vSimpleTimes{},
while Bob relocates~\vSimpleTimes{} under~\vMultiCycleTimes{}. 
This results in \autoref{fig:Disconnect:b} and \autoref{fig:Disconnect:c}, respectively.
Merging these causes both vertices to become disconnected, because they were both relocated. The inserted edges form a \emph{unicycle}, meaning a cycle where every vertex has in-degree $1$. In this case, we cannot rely on relocation references to break the cycle. Instead, we break the cycle by arbitrarily but deterministically choosing an edge along the unicycle, e.g. the edge with the smallest UID, and leaving a \emph{unicycle conflict reference} at that location, as shown in \autoref{fig:Disconnect:d}. A user can be notified of this situation when merging and resolve these conflicts again by relocating or deleting terms until the cycle no longer exists in the graph.

Relocation conflict references and unicycle conflict references together fully address the \textbf{relocation conflict problem} from \autoref{sec:Introduction}.

@@
